Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (\emph{Sida fallax})

### 2023-11-01 ms/ms.qmd
slightly revised habitat-climate results based on updated analysis

\begin{figure}[H]
  \includegraphics[height=3.25in]{../figures/study-system.pdf}
  \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
  \caption{A. Typical growth form of montane (left) and coastal (right) ‘ilima plants and B. leaves. C. Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island). D. Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study. Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error. The habitats did not differ significantly in mean annual precipitation (top-left) or solar radiation (bottom-left), but montane plants inhabit significantly cooler climates (top-right) with greater vegetation height (bottom-right). MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
  \label{fig:study-system}
\end{figure}

@@

Coastal ʻilima are surrounded by shorter vegetation than their montane counterparts (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Veg'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Veg'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Veg'))[['p']], 3)`$). The montane site with the lowest vegetation height is a remnant dry forest (Koaiʻa tree sanctuary) in a matrix of cattle pasture, hence the satellite derived vegetation height may be lower than what existed prior to human disturbance. Habitats did not differ significantly in average solar radiation at the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Sol'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Sol'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Sol'))[['p']], 3)`$). Coastal habitats are significantly warmer (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAT'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAT'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAT'))[['p']], 3)`$) but mean annual precipitation is similar (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAP'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAP'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAP'))[['p']], 3)`$).

@@

Three limitations of this study are the small sample size, experimental design that precludes distinguishing genetic from environmental differences in leaf traits, and potentially confounding effects of other environmental differences besides light environment. Understanding the mechanistic basis of higher $\mathrm{AA}$ in sun leaves would require much larger sample sizes. Sun leaves tend to be thicker, more densely packed with mesophyll cells, and have greater photosynthetic capacity and higher stomatal conductance, among other traits [@lambers_plant_2008]. Each of these factors and others potentially modulate $\mathrm{AA}$. Quantifying the contribution of all these factors requires larger samples and additional measurements that are beyond the scope of this study, but exciting avenues for future research on leaf structure-function relations. Although many morphological traits that distinguish coastal and montane ʻilima populations persist in a common environment [@yorkston_interfertility_2006], we cannot distinguish between genetic effects and plastic responses to habitat as causes of difference in $\mathrm{AA}$ because we measured naturally occurring plants *in situ*. While disentangling genetic and plastic contributions is not necessarily important for understanding the distribution of amphistomatous leaves, it would be insightful to know about genetic and environmental contributions to trait variation. A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature. However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature. Experimental studies in controlled environments will be necessary to isolate the effect light quantity and quality on $\mathrm{AA}$.

>>>>

\begin{figure}[H]
  \includegraphics[height=3.25in]{../figures/study-system.pdf}
  \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
  \caption{A. Typical growth form of montane (left) and coastal (right) ‘ilima plants and B. leaves. C. Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island). D. Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study. Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error. The habitats differ significantly in mean annual precipitation (top-left), solar radiation (bottom-left), temperature (top-right), and vegetation height (bottom-right). MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
  \label{fig:study-system}
\end{figure}

@@

Coastal ʻilima are surrounded by shorter vegetation than their montane counterparts (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Veg'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Veg'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Veg'))[['p']], 3)`$). The montane site with the lowest vegetation height is a remnant dry forest (Koaiʻa tree sanctuary) in a matrix of cattle pasture, hence the satellite derived vegetation height may be lower than what existed prior to human disturbance. Coastal sites receive greater average solar radiation at the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Sol'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Sol'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Sol'))[['p']], 3)`$); coastal sites are significantly warmer (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAT'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAT'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAT'))[['p']], 3)`$); and coastal sites receive less precipitation (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAP'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAP'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAP'))[['p']], 3)`$).

@@

Three limitations of this study are the small sample size, experimental design that precludes distinguishing genetic from environmental differences in leaf traits, and potentially confounding effects of other environmental differences besides light environment. Understanding the mechanistic basis of higher $\mathrm{AA}$ in sun leaves would require much larger sample sizes. Sun leaves tend to be thicker, more densely packed with mesophyll cells, and have greater photosynthetic capacity and higher stomatal conductance, among other traits [@lambers_plant_2008]. Each of these factors and others potentially modulate $\mathrm{AA}$. Quantifying the contribution of all these factors requires larger samples and additional measurements that are beyond the scope of this study, but exciting avenues for future research on leaf structure-function relations. Although many morphological traits that distinguish coastal and montane ʻilima populations persist in a common environment [@yorkston_interfertility_2006], we cannot distinguish between genetic effects and plastic responses to habitat as causes of difference in $\mathrm{AA}$ because we measured naturally occurring plants *in situ*. While disentangling genetic and plastic contributions is not necessarily important for understanding the distribution of amphistomatous leaves, it would be insightful to know about genetic and environmental contributions to trait variation. A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature. However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature and precipitaiton. Experimental studies in controlled environments will be necessary to isolate the effect light quantity and quality on $\mathrm{AA}$.

### 2023-11-01 ms/ms.tex
slightly revised habitat-climate results based on updated analysis

% toggle line numbers
\usepackage{lineno}
% \linenumbers

% toggle spacing
\usepackage{setspace}
\singlespacing
% \doublespacing

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

@@

Coastal ʻilima are surrounded by shorter vegetation than their montane
counterparts (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{5.41} = 4.29\), \(P = 0.007\)). The montane site
with the lowest vegetation height is a remnant dry forest (Koaiʻa tree
sanctuary) in a matrix of cattle pasture, hence the satellite derived
vegetation height may be lower than what existed prior to human
disturbance. Habitats did not differ significantly in average solar
radiation at the top of the canopy (\autoref{fig:study-system}d; Welch
Two Sample \emph{t}-test, \(t_{9.93} = -1.89\), \(P = 0.089\)). Coastal
habitats are significantly warmer (\autoref{fig:study-system}d; Welch
Two Sample \emph{t}-test, \(t_{5.01} = -2.86\), \(P = 0.035\)) but mean
annual precipitation is similar (\autoref{fig:study-system}d; Welch Two
Sample \emph{t}-test, \(t_{6.15} = 2.21\), \(P = 0.068\)).

@@

Three limitations of this study are the small sample size, experimental
design that precludes distinguishing genetic from environmental
differences in leaf traits, and potentially confounding effects of other
environmental differences besides light environment. Understanding the
mechanistic basis of higher \(\mathrm{AA}\) in sun leaves would require
much larger sample sizes. Sun leaves tend to be thicker, more densely
packed with mesophyll cells, and have greater photosynthetic capacity
and higher stomatal conductance, among other traits (Lambers et al.,
2008). Each of these factors and others potentially modulate
\(\mathrm{AA}\). Quantifying the contribution of all these factors
requires larger samples and additional measurements that are beyond the
scope of this study, but exciting avenues for future research on leaf
structure-function relations. Although many morphological traits that
distinguish coastal and montane ʻilima populations persist in a common
environment (Yorkston and Daehler, 2006), we cannot distinguish between
genetic effects and plastic responses to habitat as causes of difference
in \(\mathrm{AA}\) because we measured naturally occurring plants
\emph{in situ}. While disentangling genetic and plastic contributions is
not necessarily important for understanding the distribution of
amphistomatous leaves, it would be insightful to know about genetic and
environmental contributions to trait variation. A reciprocal transplant
would be able to determine the genetic and environmental contributions,
as well their interaction, to trait variance in nature. However,
reciprocal transplants cannot control for other differences between
coastal and montane habitats besides vegetation height, such as
temperature. Experimental studies in controlled environments will be
necessary to isolate the effect light quantity and quality on
\(\mathrm{AA}\).

>>>>

% toggle line numbers
\usepackage{lineno}
\linenumbers

% toggle spacing
\usepackage{setspace}
% \singlespacing
\doublespacing

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

Coastal ʻilima are surrounded by shorter vegetation than their montane
counterparts (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{5.41} = 4.29\), \(P = 0.007\)). The montane site
with the lowest vegetation height is a remnant dry forest (Koaiʻa tree
sanctuary) in a matrix of cattle pasture, hence the satellite derived
vegetation height may be lower than what existed prior to human
disturbance. Coastal sites receive greater average solar radiation at
the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{9.93} = -1.89\), \(P = 0.089\)); coastal sites are
significantly warmer (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{5.01} = -2.86\), \(P = 0.035\)); and coastal sites
receive less precipitation (\autoref{fig:study-system}d; Welch Two
Sample \emph{t}-test, \(t_{6.15} = 2.21\), \(P = 0.068\)).

@@

Three limitations of this study are the small sample size, experimental
design that precludes distinguishing genetic from environmental
differences in leaf traits, and potentially confounding effects of other
environmental differences besides light environment. Understanding the
mechanistic basis of higher \(\mathrm{AA}\) in sun leaves would require
much larger sample sizes. Sun leaves tend to be thicker, more densely
packed with mesophyll cells, and have greater photosynthetic capacity
and higher stomatal conductance, among other traits (Lambers et al.,
2008). Each of these factors and others potentially modulate
\(\mathrm{AA}\). Quantifying the contribution of all these factors
requires larger samples and additional measurements that are beyond the
scope of this study, but exciting avenues for future research on leaf
structure-function relations. Although many morphological traits that
distinguish coastal and montane ʻilima populations persist in a common
environment (Yorkston and Daehler, 2006), we cannot distinguish between
genetic effects and plastic responses to habitat as causes of difference
in \(\mathrm{AA}\) because we measured naturally occurring plants
\emph{in situ}. While disentangling genetic and plastic contributions is
not necessarily important for understanding the distribution of
amphistomatous leaves, it would be insightful to know about genetic and
environmental contributions to trait variation. A reciprocal transplant
would be able to determine the genetic and environmental contributions,
as well their interaction, to trait variance in nature. However,
reciprocal transplants cannot control for other differences between
coastal and montane habitats besides vegetation height, such as
temperature and precipitaiton. Experimental studies in controlled
environments will be necessary to isolate the effect light quantity and
quality on \(\mathrm{AA}\).
